---
title: "data preparation script"
author: "olawal"
date: "2023-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(plotly)
library(data.table)
library(dplyr)
library(ggplot2)
```

# Attributes for both student-mat.csv (Math course) and student-por.csv (Portuguese language course) datasets:
1 school - student's school (binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira)
2 sex - student's sex (binary: "F" - female or "M" - male)
3 age - student's age (numeric: from 15 to 22)
4 address - student's home address type (binary: "U" - urban or "R" - rural)
5 famsize - family size (binary: "LE3" - less or equal to 3 or "GT3" - greater than 3)
6 Pstatus - parent's cohabitation status (binary: "T" - living together or "A" - apart)
7 Medu - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8 Fedu - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9 Mjob - mother's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
10 Fjob - father's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
11 reason - reason to choose this school (nominal: close to "home", school "reputation", "course" preference or "other")
12 guardian - student's guardian (nominal: "mother", "father" or "other")
13 traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14 studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15 failures - number of past class failures (numeric: n if 1<=n<3, else 4)
16 schoolsup - extra educational support (binary: yes or no)
17 famsup - family educational support (binary: yes or no)
18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19 activities - extra-curricular activities (binary: yes or no)
20 nursery - attended nursery school (binary: yes or no)
21 higher - wants to take higher education (binary: yes or no)
22 internet - Internet access at home (binary: yes or no)
23 romantic - with a romantic relationship (binary: yes or no)
24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)
27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29 health - current health status (numeric: from 1 - very bad to 5 - very good)
30 absences - number of school absences (numeric: from 0 to 93)

# these grades are related with the course subject, Math or Portuguese:
31 G1 - first period grade (numeric: from 0 to 20)
31 G2 - second period grade (numeric: from 0 to 20)
32 G3 - final grade (numeric: from 0 to 20, output target)

Additional note: there are several (382) students that belong to both datasets . 
These students can be identified by searching for identical attributes
that characterize each student, as shown in the annexed R file.

Import data
```{r}
stud_perf <- fread("data/student_performance.csv")
stud_perf_subjects <- read.csv("data/StudentsPerformance.csv")
```

Replacement
```{r}

stud_perf <- stud_perf %>%
  mutate(goout=case_when(
    goout == 1 ~ "Very low",
    goout == 2 ~ "Low",
    goout == 3 ~ "Medium",
    goout == 4 ~ "High",
    goout == 5 ~ "Very High"
  ),
  studytime = case_when(
    studytime == 1 ~ "<2hours",
    studytime == 2 ~ "2 - 5 hours",
    studytime == 3 ~ "5 - 10 hours",
    studytime == 4 ~ ">10 hours"
  ))

fact_cols <- c("goout","studytime")
stud_perf$goout <- factor(stud_perf$goout, levels = c("Very low","Low","Medium","High","Very High"), ordered = TRUE)
stud_perf$studytime <- factor(stud_perf$studytime, levels = c("<2hours","2 - 5 hours","5 - 10 hours",">10 hours"), ordered = TRUE) 

```


Going out and study time relationship
```{r}
stud_perf[,.(mean_score=mean(G3)),by=c("goout","studytime")] %>%
  ggplot(aes(x=goout,y=studytime,fill=mean_score)) +
  geom_tile() +
  geom_text(aes(label=round(mean_score,1))) + # this is the function that helps us to label the cells
  scale_fill_gradient(low = "white",high = "blue") + # you can use your own colors using this function
  labs(title="How studytime relates to Going out",y="Study time",x="Going out") # Customizing your axis labels and title
```
Race/Ethnicity, how does it affect performance
```{r}
stud_perf_subjects$score <-rowMeans(stud_perf_subjects[c('math.score', 'reading.score','writing.score')], na.rm = TRUE)

setDT(stud_perf_subjects)[,.(mean_score=mean(score)),by=c("race.ethnicity")] %>%
plot_ly() %>%
        add_trace(x = ~reorder(race.ethnicity, mean_score), 
                  y = ~mean_score,
                  size = ~mean_score,
                  color = ~race.ethnicity,
                  alpha = 1.5,
                  #sizes = c(200,4000),
                  text = ~paste("Ethnic group: ",race.ethnicity, "\n","Mean score: ",round(mean_score,2)),
                  type = "scatter",
                  mode = "markers",
                  marker = list(symbol = 'circle', sizemode = 'diameter',
                                line = list(width = 2, color = '#FFFFFF'), opacity=0.4)) %>%
        # add_text(x = ~reorder(race.ethnicity, -mean_score), 
        #                 y = ~race.ethnicity, text = ~mean_score,
        #                 showarrow = FALSE,
        #                 color = I("black")) %>%
        layout(
          showlegend = FALSE,
          title="Mean score with respect to ethnic groups",
          xaxis = list(
            title = "Race/Ethnic group"
          ),
          yaxis = list(
            title = "Mean score"
          )
        ) %>%
        config(displayModeBar = FALSE, displaylogo = FALSE, 
              scrollZoom = FALSE, showAxisDragHandles = TRUE, 
              showSendToCloud = FALSE)
```

## Performance in regards to gender 

```{r}
# generate student id's 
stud_perf_subjects <- stud_perf_subjects[, student_id := .I]

stud_perf_subjects_melt <- melt(stud_perf_subjects,variable.name = "subjects",measure.vars = c("math.score","reading.score","writing.score"),value.name = "score")

stud_perf_subjects_melt[,.(mean_score=mean(score.1)),by=c("gender","subjects")] %>% 
  plot_ly(
                  x = ~gender,
                  y = ~mean_score,
                  color = ~subjects,
                  text = ~paste0("Gender: ", gender, "\n", "Subject: ", subjects,"\n","Mean score: ",round(mean_score,2))) %>%
               add_bars(showlegend=TRUE, hoverinfo='text' ) %>%
  layout(title="Mean gender performance",
         yaxis = list(title = "Mean Score"),
         xaxis = list(title = "Gender"))



```

Scatter plot
```{r}
stud_perf_subjects %>%
  ggplot(aes(x=math.score,y=reading.score,color=gender)) +
  geom_jitter() +
  labs(title="Reading and Math scores scatter plot",y="Reading score",x="Math score")
```

Gender mean score
```{r}

gauge.df <- stud_perf_subjects %>% group_by(gender) %>% summarise(mean.gender.score = round(mean(score),2))

 plot_ly(
    domain = list(x = c(0, 1), y = c(0, 1)),
    value = 69.57,#gauge.df[gauge.df$gender=="female"],
    title = list(text = "Female mean score"),
    type = "indicator",
    mode = "gauge+number") %>%
  layout(margin = list(l=20,r=30))


 plot_ly(
    domain = list(x = c(0, 1), y = c(0, 1)),
    value = 65.84,
    title = list(text = "Male mean score"),
    type = "indicator",
    mode = "gauge+number") %>%
  layout(margin = list(l=20,r=30))
```


Line graph and barplot comparison  
Filter for course
```{r}
stud_perf_subjects_melt <- melt(stud_perf_subjects_melt, variable.name="groupings",
                                measure.vars = c("race.ethnicity","parental.level.of.education","test.preparation.course","lunch"),
                                value.name = "group.values")

stud_perf_subjects_melt <- stud_perf_subjects_melt[,mean.score:=mean(score),by=.(group.values)]
```

Visualize 

```{r}

 stud_perf.groupings <- stud_perf_subjects_melt %>% 
  group_by(groupings,group.values) %>%
  mutate(mean.score=mean(score)) %>% ungroup()

scaleFactor <- (stud_perf.groupings$mean.score/stud_perf.groupings$mean.score)*stud_perf.groupings$nstuds

  ggplot(data = stud_perf.groupings,aes(x = group.values, y = mean.score, fill=mean.score)) +
  geom_col() +
     theme_bw() +
    theme(axis.text.x = element_text(vjust = 0.9, hjust = 0.9, angle = 45)) +
   # geom_line(data = stud_perf_subjects_melt,aes(x=group.values, y=mean.score/10),stat="identity",color="red",size=2, group=1)+
   # scale_y_continuous(sec.axis=sec_axis(~.*10,name="Number of Students")) +
  facet_wrap(~groupings, scales = "free") 

  
stud_perf.groupings %>% 
  filter(groupings == "race.ethnicity") %>% 
  ggplot() + 
  geom_col(aes(x = group.values, y = 3*mean.score)) +
  geom_line(aes(x = group.values, y = nstuds), size = 1.5, color="red", group = 1) + 
  scale_y_continuous(sec.axis = sec_axis(~., name = "Number of students")) #+
  #facet_wrap(~groupings, scales = "free")

```

```{r}  

stud.number <- stud_perf_subjects %>%
  group_by(race.ethnicity)  %>%
  mutate(number.studs = n())

stud.summary <- stud.number %>% 
  group_by(race.ethnicity) %>% 
  summarise(score = mean(math.score),number.studs=number.studs)

stud.summary <- stud.summary %>% unique()

ggplot(stud.summary) + 
  geom_col(aes(x = race.ethnicity, y = score, fill=score)) +
  geom_line(aes(x = race.ethnicity, y = 1*number.studs), size = 1.5, color="green", group = 1, stat="identity") + 
  scale_y_continuous(sec.axis = sec_axis(~./1, name = "Number of students")) +
  labs(title="Average score per ethnic group and number of students",x="Race/Ethnicity",y="Average score")

```

```{r}
stud_perf_subjects %>% group_by(gender) %>% summarise(count=n())
```


```{r}

library(plotrix)

example(twoord.plot)  

groupvalues <- stud_perf.groupings$group.values
meanscore <- stud_perf.groupings$mean.score
studs <- stud_perf.groupings$nstuds


twoord.plot(
            lx=as.character(groupvalues),
            ly=meanscore,
            rx=as.character(groupvalues),
            ry=studs,
            xlab="Group values",
  lylim=range(meanscore),rylim=range(studs),xlim = c("group A","group B","group C","group D","group E"),
  type=c("bar","l"),ylab="Mean score",#ylab.at=5,
  rylab="Number of students",#rylab.at=65,
  main="Bars on left axis, lines on right axis",
  lytickpos=50:70,rytickpos=seq(30000,1000000,by=50000),
  lcol=3,rcol=4,do.first="plot_bg()")


 going_up<-seq(3,7,by=0.5)+rnorm(9)
 going_down<-rev(60:74)+rnorm(15)
 
 twoord.plot(2:10,going_up,1:15,going_down,xlab="Sequence",
  lylim=range(going_up)+c(-1,10),rylim=range(going_down)+c(-10,2),
  type=c("bar","l"),ylab="Ascending values",ylab.at=5,
  rylab="Descending values",rylab.at=65,
  main="Bars on left axis, lines on right axis",
  lytickpos=3:7,rytickpos=seq(55,75,by=5),
  lcol=3,rcol=4,do.first="plot_bg()")
 
```

